---
title: "Workflow RNA-seq Analysis"
author: "Mariana Mescouto Lopes"
date: '2026'
subtitle: "Analysis of count matrix data"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 10
    fig_height: 6
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = FALSE
)

#Set working directory
setwd("~/Library/CloudStorage/OneDrive-Personal/Documentos/GitHub")

# Create directories for output organization
dir.create("results", showWarnings = FALSE)
dir.create("results/figures", showWarnings = FALSE)
dir.create("results/tables", showWarnings = FALSE)
dir.create("results/gsea", showWarnings = FALSE)
```

## Overview

This workflow demonstrates a basic RNA-seq analysis pipeline.  
To run the analysis, update the file paths according to your local environment.

This workflow performs:
- Data preprocessing
- Differential expression analysis
- Data visualization
- Functional enrichment analysis (GSEA)

# 1. Packges

To install packages, use: install.packages("package_name")

Then load them using: library(package_name)

```{r libraries}
suppressPackageStartupMessages({
  # Diferential analysis
  library(DESeq2)
  library(edgeR) 
  library(limma)
  library(fdrtool)
  
  # Data manipulation
  library(dplyr)
  library(readr)
  library(readxl)
  library(tibble)
  
  # Visualization
  library(ggplot2)
  library(pheatmap)
  library(ggrepel)
  library(RColorBrewer)
  library(corrplot)
  library(plotly)
  
  # Functional analysis
  library(fgsea)
  library(biomaRt)
  
  # Tables
  library(DT)
  library(kableExtra)
})
```

# 2. Data preprocessing

Load count matrix and sample metadata

```{r load-and-filter}

df <- read_delim("rnaseq_data.csv", delim = ";", show_col_types = FALSE)
metadata <- read_excel("samples.xlsx")

# Dataset information
cat(sprintf("Raw dataset: %d genes × %d columns\n", nrow(df), ncol(df)))
cat(sprintf("Metadata: %d samples\n", nrow(metadata)))
```

## Sample selection and gene formatting

```{r gene-sample}
ctrl_samples <- grep("CON_", colnames(df), value = TRUE)
treat_samples <- grep("BET_", colnames(df), value = TRUE)

cat(sprintf("Samples: %d control + %d treated\n", 
            length(ctrl_samples), length(treat_samples)))

# Set gene IDs as row names
genes <- df$gene
df <- as.matrix(df[, c(ctrl_samples, treat_samples)])
rownames(df) <- genes
```

## Filtering low expressed genes

Criteria: keep genes with ≥5 counts in at least 6 samples

```{r filter}
filter_mask <- rowSums(df >= 5) >= 6
expr_genes <- genes[filter_mask]
df_filtered <- df[filter_mask, ]

cat(sprintf("Initial number of genes: %d\n", nrow(df)))
cat(sprintf("Filtered dataset: %d genes (%.1f%% retained)\n", 
            nrow(df_filtered), 
            100 * nrow(df_filtered) / nrow(df)))
```

## Metadata alignment

```{r arrange}
metadata_filtered <- metadata %>%
  filter(ID %in% c(ctrl_samples, treat_samples)) %>%
  arrange(match(ID, c(ctrl_samples, treat_samples)))
```

Save filtered data 

```{r save}
write_csv(as.data.frame(df_filtered), "results/tables/filtered_data.csv")
```


# 3. Exploratory Data Analysis

## Data distribution

Evaluate data quality and filtering criteria

```{r basic-observations}
total_expression <- rowSums(df_filtered)

p_dist <- ggplot(data.frame(total_expression), 
                 aes(x = log10(total_expression + 1))) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = log10(c(10, 100, 1000)), 
             linetype = "dashed", color = "red", alpha = 0.7) +
  labs(
    title = "Expression distribution",
    subtitle = "Red lines: 10, 100, 1000 reads",
    x = "log10(total expression)",
    y = "Number of genes"
  ) +
  theme_minimal()

print(p_dist)
```

Most genes show low expression levels, which is expected in RNA-seq data. The filtering step removes lowly expressed genes, reducing noise and improving downstream analysis.

## PCA

Explore sample variability and identify potential clustering patterns

1) Normalization 

```{r normal}
df_log <- log2(df_filtered + 1)
```

2) PCA computation

```{r pca}
pca <- prcomp(t(df_log), scale. = TRUE)
pca_data <- data.frame(pca$x)

pca_data$sample <- rownames(pca_data)
pca_data$group <- ifelse(grepl("CON_", pca_data$sample), "Control", "Treated")
```

3) PCA plot

```{r plot}
p_pca <- ggplot(pca_data, aes(PC1, PC2, color = group)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "PCA of RNA-seq samples",
    x = paste0("PC1 (", round(summary(pca)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca)$importance[2,2]*100, 1), "%)")
  ) +
  theme_minimal()

print(p_pca)
```

PCA revealed no clear separation between control and treated samples, suggesting that the treatment effect is relatively subtle compared to the overall variability in the dataset.

This may reflect either a mild biological response or inherent variability between samples.

## Hierarchical clustering

Assess sample similarity.

```{r cluster}
dist_mat <- dist(t(df_log))
hc <- hclust(dist_mat)

plot(hc, main = "Hierarchical clustering of samples")
```
Hierarchical clustering did not fully separate samples by condition, further supporting the observation that the treatment effect is modest.

Some samples cluster independently of their group, indicating variability within conditions.

## Heatmap

Top 10 most highly expressed genes

```{r heatmap-10}
top_genes <- head(order(rowMeans(df_log), decreasing = TRUE), 10)

pheatmap(df_log[top_genes, ],
         scale = "row",
         main = "Top 10 most expressed genes")
```
The heatmap of highly expressed genes shows no strong clustering by condition, suggesting that the most abundant genes are not strongly affected by the treatment.

This is consistent with the PCA and clustering results.

# 4. Differential analysis

Organize metadata

```{r diff-analysis}
coldata <- DataFrame(
  treatment = factor(metadata_filtered$TRAT),
  ID = metadata_filtered$Identificação,
  row.names = metadata_filtered$ID
)

#Check order
if (!identical(colnames(df), rownames(coldata))) {
  cat("⚠️Sampless...\n")
  coldata <- coldata[colnames(df_filtered), ]
}
```

DESeq2

```{r deseq2}
dds <- DESeqDataSetFromMatrix(
  countData = df_filtered,
  colData = coldata, 
  design = ~ treatment
)

dds <- DESeq(dds)
```

Results

```{r results-deseq}
res <- results(dds, contrast = c("treatment", "BET", "CON"))

res_summary <- data.frame(
  Info = c("p-adj < 0.05", "p-adj < 0.10", "p-adj < 0.20"),
  Value = c(
    sum(res$padj < 0.05, na.rm = TRUE),
    sum(res$padj < 0.10, na.rm = TRUE),
    sum(res$padj < 0.20, na.rm = TRUE)
  )
)

kable(res_summary) %>% kable_styling()
```
